---
title: "Introduction to non-compartmental analysis using R"
output:
  bookdown::html_document2:
    fig_caption : TRUE
    number_sections: no
    toc: yes
    toc_float: yes

header-includes:
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \captionsetup{labelfont=bf,margin=12pt} 
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

<style>
h1.title {
  font-size: 48px;

  text-align: center;
}
h1 {
  font-size: 24px;
}
h2 {
  font-size: 20px;
}
h3 {
  font-size: 16px;
}

h4.author { 
  text-align: center;
}
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.pos= "H")
library(tidyverse)
library(kableExtra)
library(PKNCA)
```

# Learning outcomes  

By the end of this session you should be able to:  

* Interpret a concentration-time plot
  
* Use  <TT>R</TT> to undertake basic, descriptive (non-compartmental) analysis of some pharmacokinetic data to estimate
    + $C_{max}$
    + $T_{max}$
    + $AUC_{0-t}$

* Describe the relationship between clearance, dose and $AUC$

* List some benefits and limitations of non-compartmental PK analysis    

# Tasks

* Make a concentration-time plot for the 'Healthy' dataset
* Make a concentration-time plot for 'Sepsis' dataset

* Calculate $C_{max}$, $T_{max}$ and $AUC_{0-t}$ for Healthy dataset
* Calculate $C_{max}$, $T_{max}$ and $AUC_{0-t}$ for Sepsis dataset

* Estimate Vd and Cl for Healthy dataset
* Estimate Vd and Cl for Sepsis dataset

Extension tasks:

* Can you put plots of Healthy and Sepsis datasets on the same Figure?
* Can you make a table to compare $C_{max}$, $T_{max}$ and $AUC_{0-t}$ for Healthy and Sepsis datasets?

# Packages you need for this session

Link your path using the following line of code and then call <TT>library(tidyverse)</TT>.

```{r, eval=FALSE}
.libPaths( c( .libPaths(),
              "/homes/dlonsdale-pharmacokinetics/sghms/bms/shares/Advanced-Pharmacokinetics/4.4.2/library") )
#note that you need the '.' before 'libPaths'
library(tidyverse)
```

# Teaching data for this session

For this session we will be using PK data for an anti-epileptic drug. 6 participants were given an IV dose and then one week later they were given an oral dose. Drug concentrations were measured.

If you want to access these datasets yourself you can copy the code below. 

```{r, eval=TRUE}

IV<-read.csv("https://raw.githubusercontent.com/dlonsdal/non_compartmental/main/docs/PK_IV.csv")


ORAL<-read.csv("https://raw.githubusercontent.com/dlonsdal/non_compartmental/main/docs/PK_oral.csv")

```

**Task:** Plot the data for the IV and oral data.

More detailed notes on putting together plots are included in the Introduction to R page.  

```{r plot, eval=FALSE,echo=FALSE, message=FALSE, warning=FALSE}


ggplot(data=IV,aes(x=time, y=conc,col=id) )+
  geom_line()+
  geom_point()+
  theme_classic()+
  xlab("time (hours)")+
  ylab("concentration of IV drug (mg/L)")

```

# Estimating $C_{max}$

Remember that $C_{max}$ is observed from our data. We find the max value per participant and report it. We actually did this in the last session. 

**Tasks:** 

* Estimate $C_{max}$ for the IV and oral data
* Estimate the mean $C_{max}$ for the IV and oral data

_Advanced: Output both IV and oral data together in one table_

```{r cmax_solution, eval=FALSE,echo=FALSE, message=FALSE, warning=FALSE}
IV%>%
  group_by(id)%>%
  summarise(cmax=max(conc))
# here we tell R to take 'IV', group by 'id' then give the max conc

ORAL %>% 
  group_by(id) %>% 
  summarise(cmax=max(conc))

IV%>%
  group_by(id)%>%
  summarise(cmax=max(conc)) %>% 
  ungroup() %>% 
  summarise(mean(cmax))
# here we tell R to take 'IV', group by 'id' then give the max conc

ORAL %>% 
  group_by(id) %>% 
  summarise(cmax=max(conc)) %>% 
  ungroup() %>% 
  summarise(mean(cmax))

FULL<-rbind(IV,ORAL)
FULL %>% 
  group_by(id,study) %>% 
  summarise(max(conc))

FULL %>% 
  group_by(id,study) %>% 
  summarise(cmax=max(conc),.groups="drop") %>% 
  group_by(study)  %>% 
  summarise(mean(cmax))

```

# Estimating $T_{max}$

This is a little more tricky. We need to get <TT>R</TT> to find $C_{max}$ and then keep the associated value for time.

Try a Google search for this. Something like 'using group by to get the value corresponding to the max value of another column'.

**Task:** Estimate $T_{max}$ for the IV and oral data

```{r tmax_solution,  eval=FALSE,echo=FALSE, message=FALSE, warning=FALSE}

ORAL %>% 
  group_by(id) %>% 
  summarise(cmax=max(conc),
            tmax=time[which.max(conc)])

FULL %>% 
  group_by(id,study) %>% 
  summarise(cmax=max(conc),
            tmax=time[which.max(conc)],.groups="drop") %>% 
  group_by(study)  %>% 
  summarise(mean(cmax),
            mean(tmax))  
```

# Estimating the area under the curve ($AUC_{0-t}$)

$AUC_{0-t}$ is the area under the time concentration curve from $time=0$ to $time=t$ ($t$ will usually be the time of the last observed sample but it doesnt have to be). In non-compartmental analysis, this is often estimated using a method that breaks up the time concentration data into small chunks, estimates the area under each chunk and adds them together. The most straightforward way of doing this is using the trapezoid method. 

```{r trap, echo=F, fig.cap="Participant 'A' oral concentration-time profile, with trapezoids indicated in colour", fig.width=7,fig.height=2.5,fig.align = 'center', warning=F,message=F}

soral<-subset(ORAL,id=='A')
a<-ggplot(data=soral, aes(time,conc))+
  geom_point()+
  xlab("Time (hours)") + ylab("Drug concentration (mg/L)")+
  geom_line() +
  theme_bw(base_size=9)

b<-ggplot(data=soral, aes(time,conc))+ 

    geom_ribbon(data=subset(soral,  time >2 & time < 7), 
          aes(ymin=0,ymax=conc),
              fill='#82B446',
              alpha=0.2, outline.type = 'full',colour='gray44')+ 

  geom_point()+
  xlab("Time (hours)") + ylab("Drug concentration (mg/L)")+
  geom_line() +
  theme_bw(base_size=9) +
  geom_text(x=4, y=0, label="A", hjust= -0.25, vjust = 1, size=3)+
  geom_text(x=4, y=soral$conc[9], label="B",hjust= -0.5,  size=3)+
  geom_text(x=6, y=soral$conc[10], label="C", hjust= -0.5,vjust=-0.25, size=3)+
  geom_text(x=6, y=0, label="D", hjust= -0.25, vjust = 1, size=3)

library(gridExtra)
grid.arrange(a,b,ncol=2)
```

To calculate the area under the curve of the trapezoid ABCD in Figure \@ref(fig:trap) we use the equation for the area of a trapezoid. $$Area = \frac{1}{2}*(height_{1,AB}+height_{2,CD})*base_{AD}$$
The two heights correspond to concentrations B and C and the base is the distance A to D (the change in time $\Delta t$) in Figure \@ref(fig:trap). So more fully for this example, where *height* is replaced with *concentration* $(C)$: $$AUC_{0-8} = \sum_{t=0}^{t=8} \Delta t\frac{(C_1+C_2)}{2}$$ 

Fortunately,<TT>R</TT> has an inbuilt function that allows us to calculate the $AUC$. The handily named <TT>AUC()</TT> function. You need to provide it with three commands, <TT>x=...,y=...</TT> and <TT>method="trapezoid"</TT>. It is in the <TT>DescTools</TT> library.

I will show you how to calculate using the oral data. 

```{r AUC,   message=FALSE, warning=FALSE}
library(DescTools)
ORAL%>% 
  group_by(id) %>% 
  summarise(AUC=AUC(time,conc,method='trapezoid'))

```

**Tasks:**

* Calculate $AUC$ for the IV data
* Calculate mean $AUC$ for each dataset
* Estimate the bioavailability ($F$) of the drug

```{r AUC_solutions,  eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
IV%>% 
  group_by(id) %>% 
  summarise(AUC=AUC(time,conc,method='trapezoid'))

FULL %>% 
  group_by(id,study) %>% 
  summarise(AUC=AUC(time,conc,method='trapezoid')) %>% 
  group_by(study)  %>% 
  summarise(mAUC=
              mean(AUC)) 

FULL %>% 
  group_by(id,study) %>% 
  summarise(AUC=AUC(time,conc,method='trapezoid')) %>% 
  group_by(study)  %>% 
  summarise(mAUC=
              mean(AUC)) %>% 
  summarise(F=mAUC[study=="ORAL"]/mAUC[study=="IV"])


```

## Limitations of the linear trapezoid method 

```{r trapbest, echo=F, fig.cap="Participant 'A' IV concentration-time profile with curved line of best fit", fig.width=4.5,fig.height=2.5,fig.align = 'center'}
sIV<-subset(IV,id=="A"&time>4)
ggplot(data=sIV, aes(time,conc))+ 
    geom_ribbon(data=subset(sIV,  time >11 & time < 25), 
          aes(ymin=0,ymax=conc),
              fill='#82B446',
              alpha=0.2, outline.type = 'full',colour='gray44')+ 

  geom_point()+
  xlab("Time (hours)") + ylab("Drug concentration (mg/L)")+
  geom_line()+ 

  geom_line()  +
  geom_function(fun=function(x) (1000/27.75)*exp((-0.17)*x),linetype='twodash', col='red')+
  labs(caption = "Line of best fit shown in red")

```

It is worth noting the trapezoid method has limitations. Most notably that it will *overestimate* $AUC$ by virtue of taking a straight line between points, when the true concentration will fall in an logarithmic (curve) fashion (Figure \@ref(fig:trapbest)). There are methods to improve this accuracy, such as the logarithmic trapezoidal method. With this method, the 'trapezoid' is given a curved top (like the red line in Figure \@ref(fig:trapbest)). The formula for this method is $$AUC = \sum_{t=0}^{t=\tau} \frac{\Delta t}{ln(\frac{C_1}{C_2})} (C_1+C_2)$$  

\newpage 

### Using <TT>R</TT> package <TT>PKNCA</TT> to undertake non-compartmental analysis  

If that all seems like too much hard work and <TT>R</TT> should just jolly well get on and do for us then you are right! There are a few non-compartmental analysis packages available, we will use <TT>PKNCA</TT>. Remember you will need to run <TT>install.packages("PKNCA")</TT> the first time you wish to use this package, once it has been installed you need to run <TT>library(PKNCA)</TT>. 

To run non-compartmental analysis using this package, we need to give <TT>PKNCA</TT> some dosing data. We know that all the participants were given a dose of 1000 mg of amoxicillin at time $t=0$. To add a new column (variable) using <TT>\%>\%</TT> we use the <TT>mutate</TT> function.  

```{r, results='hold'}
amox<- read.csv("https://raw.githubusercontent.com/dlonsdal/SGUL_PK_data/refs/heads/main/amoxicillinPK.csv") 

damox<-amox %>%  # remember we will need to save this as a new object
  subset(time==0) %>% 
  mutate(dose=1000) #this produces a new variable dose 1000
```

**Task:** Run this line of code and have a look at the newly created 'damox' object 

Now we can call and run <TT>PKNCA</TT>. First we create a sampling (or concentration) object and then a dosing object. In each function we instruct <TT>PKNCA</TT> of the names of our data object and the relevant variables in it. Then we combine the two and run our non-compartmental analysis. 

```{r, results='hold', eval=FALSE}
library(PKNCA)
conc<-PKNCAconc(amox, conc~time|ID) #create sampling object
dose<-PKNCAdose(damox, dose~time|ID) #create dosing object
full<-PKNCAdata(conc, dose)  #combine
results<-pk.nca(full)  # analyse
summary(results) #summarise the results
```
```{r, results='hold', echo=FALSE}
library(PKNCA)
conc<-PKNCAconc(amox, conc~time|ID) #create sampling object
dose<-PKNCAdose(damox, dose~time|ID) #create dosing object
full<-PKNCAdata(conc, dose)  #combine
results<-pk.nca(full)  # analyse

knitr::kable(summary(results), "latex",booktabs = TRUE, caption='PKNCA output')%>%
  kable_styling(latex_options = "striped")
```


**Task:** Look at the value for $AUC$ produced here and compare it to the one earlier. What do you notice? Why do you think there is a discrepancy?

The answer is that by default, PKNCA uses logarithmic trapezoid method to calculate $AUC$. If you want to set this to linear logarithmic method you can set this as an option. Note that it also uses the 'geometric' rather than arithmetic mean.  


```{r, results='hold',eval=F}
full<-PKNCAdata(conc, dose,options=list(auc.method="linear")) 
results<-pk.nca(full)  
summary(results)
```
```{r, results='hold',echo=F, eval=T}
full<-PKNCAdata(conc, dose,options=list(auc.method="linear"))  #combine
results<-pk.nca(full)  # analyse
knitr::kable(summary(results), "latex",booktabs = TRUE, caption='PKNCA output, linear logarithmic AUC method')%>%
  kable_styling(latex_options = "striped")
```

```{r,echo = FALSE,eval=FALSE}
PKNCA.set.summary(
      name="half.life",
      description="arithmetic mean and standard deviation",
      point=business.mean,
      spread=business.sd,
      rounding=list(signif=3)
    )

#not run but code present in case someone asks how to do arithmetic mean
```

## Estimating clearance
Remember that clearance can be calculated as $$CL = \frac{Dose}{AUC}$$

\newpage

## Limitations of non-compartmental analysis

```{r traperror, echo=F, fig.cap="Participant A amoxicillin concentration-time profile showing impact of reduced sampliing on AUC calculation", fig.width=4.5,fig.height=2.5,fig.align = 'center'}

Aamox<-amox%>%
        subset(ID=='A')

ggplot(data=Aamox, aes(time,conc))+ 
    geom_ribbon(data=subset(Aamox,time < 5&time!=2), 
          aes(ymin=0,ymax=conc),
              fill='#B4464B',
              alpha=0.1, outline.type = 'full',colour='gray44')+ 
    geom_ribbon(data=subset(Aamox,  time < 3), 
          aes(ymin=0,ymax=conc),
              fill='blue',
              alpha=0.1, outline.type = 'full',colour='gray44')+ 
    geom_ribbon(data=subset(Aamox,  time >1 & time < 5), 
          aes(ymin=0,ymax=conc),
              fill='#82B446',
              alpha=0.1, outline.type = 'full',colour='gray44')+
  geom_point()+
  theme_bw(base_size=9)+
  xlab("Time (hours)") + ylab("Concentration of amoxicillin (mg/L)")+
  theme_bw(base_size=9)+
  geom_line() +
  geom_function(fun=function(x) 1000/23.65*exp((-11.4/23.65)*x),col='red')+
  labs(caption = "Pink shaded area shows the increased error associated with reduced sampling \n (time = 2 sample removed)")

```

Non-compartmental analyses are useful for many reasons. For example, they use relatively straightforward algebra with few assumptions about the data to calculate summary pharmacokinetic information. Most regulators will require non-compartmental analysis of data as part of regulatory submissions. There are some disadvantages that are worth being aware of. These include:

* Need for rich sampling (multiple samples from each subject)
* Use of direct observation of C~max and T~max means *true* values are missed for any particular subject
* Increasing accuracy requires increased number of participants or samples, which can be technically chalenging and expensive

Have a look at Figure \@ref(fig:traperror), where a situation is imagined that $time=2$ data point is not sampled. You can see the likely impact on the estimate of AUC from this. 