---
title: "Introduction to non-compartmental analysis using R"
output:
  bookdown::html_document2:
    fig_caption : TRUE
    number_sections: no
    toc: yes
    toc_float: yes
header-includes:
  - \usepackage{colortbl}
  - \usepackage{multirow}
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \captionsetup{labelfont=bf,margin=12pt} 
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

<style>
h1.title {
  font-size: 48px;

  text-align: center;
}
h1 {
  font-size: 24px;
}
h2 {
  font-size: 20px;
}
h3 {
  font-size: 16px;
}

h4.author { 
  text-align: center;
}
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.pos= "H")
library(tidyverse)
library(kableExtra)
library(PKNCA)
library(gridExtra)
library(DescTools)
```


# Learning outcomes  

By the end of this session you should be able to:  

* Interpret a concentration-time plot
  
* Use  <TT>R</TT> to undertake basic, descriptive (non-compartmental) analysis of pharmacokinetic data to estimate
    + $C_{max}$
    + $T_{max}$
    + $AUC_{0-t}$

* Describe the relationship between clearance, dose and $AUC$

* List benefits and limitations of non-compartmental PK analysis    

# Tasks

* Make a concentration-time plot for the 'Healthy' dataset

* Make a concentration-time plot for 'Sepsis' dataset

* Estimate $C_{max}$ and $AUC_{0-t}$ for the 'Healthy' dataset

* Estimate $C_{max}$ and $AUC_{0-t}$ for the 'Sepsis' dataset

* Estimate $Cl$ for the 'Healthy' dataset

* Estimate $Cl$ for the 'Sepsis' dataset

### Extension tasks:

* Plot 'Healthy' and 'Sepsis' concentrationt-time profiles on the same Figure

* Make a table comparing $C_{max}$ and $AUC_{0-t}$ for 'Healthy' and 'Sepsis' datasets

# Packages you need for this session

```{r, eval=FALSE}
library(tidyverse)
library(PKNCA)
library(DescTools)
```

# Datasets for tasks during this session

Patients with and without sepsis have received a 1000 mg *intravenous bolus* of Drug A, a renally-cleared antibiotic.

```{r datasets, eval=FALSE}

healthy <- read_csv("https://raw.githubusercontent.com/c-swire/pkcourse/refs/heads/main/healthy_nca.csv")
sepsis <- read_csv("https://raw.githubusercontent.com/c-swire/pkcourse/refs/heads/main/sepsis_nca.csv")

```

# Datasets used for workbook examples

This workbook uses PK data for an anti-epileptic drug. 6 participants were given an *IV bolus* dose and then one week later they were given an *oral* dose. Serial drug concentrations were measured.

If you want to access these datasets yourself you can copy the code below. 

```{r, eval=TRUE}

IV<-read.csv("https://raw.githubusercontent.com/dlonsdal/non_compartmental/main/docs/PK_IV.csv")
ORAL<-read.csv("https://raw.githubusercontent.com/dlonsdal/non_compartmental/main/docs/PK_oral.csv")

```

The first six lines of the ORAL dataset are shown here.

``` {r ivhead, eval=TRUE}

head(ORAL)

```

Columns in these datasets are:

* id = patient ID
* time = time after dose administered
* study = whether IV or ORAL study
* conc = drug concentration

**How to:** Make a concentration-time plot for the ORAL data.

_Hint: More detailed notes on putting together plots are included in the Introduction to R page_

```{r plot,  message=FALSE, warning=FALSE}

ggplot(data=ORAL,aes(x=time, y=conc,col=id) )+
  geom_line()+
  geom_point()+
  theme_classic()+
  xlab("time (hours)")+
  ylab("concentration of drug (mg/L)")

```


# Estimating $C_{max}$

$C_{max}$ is observed from our data. We find the maximum value per participant and report it. 

**How to:** 

* Estimate $C_{max}$ for the oral data
* Estimate the mean $C_{max}$ for the oral data

```{r cmax_solution, message=FALSE, warning=FALSE}

# here we tell R to take 'ORAL', group by 'id' then give the max conc
ORAL %>% 
  group_by(id) %>% 
  summarise(cmax=max(conc))

# here we tell R to take 'ORAL', group by 'id', find the max conc for each 'id' and then output the mean of the those values
ORAL %>% 
  group_by(id) %>% 
  summarise(cmax=max(conc)) %>% 
  ungroup() %>% 
  summarise(mean(cmax))

```

# Estimating $T_{max}$

This is a little more tricky. We need to get <TT>R</TT> to find $C_{max}$ and then keep the associated value for time.

**How to:** Estimate $T_{max}$ for the oral data

```{r tmax_solution,   message=FALSE, warning=FALSE}

ORAL %>% 
  group_by(id) %>% 
  summarise(cmax=max(conc),
            tmax=time[which.max(conc)])

# we can also output the results for both datasets together

FULL<-rbind(IV,ORAL) # to combine the two datasets

FULL %>% 
  group_by(id,study) %>% 
  summarise(cmax=max(conc),
            tmax=time[which.max(conc)],.groups="drop") %>% 
  group_by(study)  %>% 
  summarise(mean(cmax),
            mean(tmax))  
```

As you can see here, $T_{max}$ after an IV bolus is at t=0. As a result, this is not a parameter that is reported for IV bolus studies.  

# Estimating the area under the curve ($AUC_{0-t}$)

$AUC_{0-t}$ is the area under the time concentration curve from $time=0$ to $time=t$ ($t$ will usually be the time of the last observed sample but it doesn't have to be). In non-compartmental analysis, this is often estimated using a method that breaks up the time concentration data into small chunks, estimates the area under each chunk and adds them together. The most straightforward way of doing this is using the trapezoid method. 

```{r trap, echo=F, fig.cap="Participant 'A' oral concentration-time profile, with trapezoids indicated in colour", fig.width=7,fig.height=2.5,fig.align = 'center', warning=F,message=F}

soral<-subset(ORAL,id=='A')
a<-ggplot(data=soral, aes(time,conc))+
  geom_point()+
  xlab("Time (hours)") + ylab("Drug concentration (mg/L)")+
  geom_line() +
  theme_bw(base_size=9)

b<-ggplot(data=soral, aes(time,conc))+ 

    geom_ribbon(data=subset(soral,  time >2 & time < 7), 
          aes(ymin=0,ymax=conc),
              fill='#82B446',
              alpha=0.2, outline.type = 'full',colour='gray44')+ 

  geom_point()+
  xlab("Time (hours)") + ylab("Drug concentration (mg/L)")+
  geom_line() +
  theme_bw(base_size=9) +
  geom_text(x=4, y=0, label="A", hjust= -0.25, vjust = 1, size=3)+
  geom_text(x=4, y=soral$conc[9], label="B",hjust= -0.5,  size=3)+
  geom_text(x=6, y=soral$conc[10], label="C", hjust= -0.5,vjust=-0.25, size=3)+
  geom_text(x=6, y=0, label="D", hjust= -0.25, vjust = 1, size=3)


grid.arrange(a,b,ncol=2)
```

To calculate the area under the curve of the trapezoid ABCD in Figure \@ref(fig:trap) we use the equation for the area of a trapezoid. $$Area = \frac{1}{2}*(height_{1,AB}+height_{2,CD})*base_{AD}$$
The two heights correspond to concentrations B and C and the base is the distance A to D (the change in time $\Delta t$) in Figure \@ref(fig:trap). So more fully for this example, where *height* is replaced with *concentration* $(C)$: $$AUC_{0-8} = \sum_{t=0}^{t=8} \Delta t\frac{(C_1+C_2)}{2}$$ 

Fortunately, <TT>R</TT> has an inbuilt function that allows us to calculate the $AUC$. The handily named <TT>AUC()</TT> function. You need to provide it with three commands, <TT>x=...,y=...</TT> and <TT>method="trapezoid"</TT>. It is in the <TT>DescTools</TT> library.

Here is how to calculate using the oral data:

```{r AUC,   message=FALSE, warning=FALSE}

ORAL%>% 
  group_by(id) %>% 
  summarise(AUC=AUC(time,conc,method='trapezoid'))

```

**How to:**

* Calculate mean $AUC$ for each dataset
* Estimate the bioavailability ($F$) of the drug

```{r AUC_solutions, message=FALSE, warning=FALSE}

FULL %>% 
  group_by(id,study) %>% 
  summarise(AUC=AUC(time,conc,method='trapezoid')) %>% 
  group_by(study)  %>% 
  summarise(mAUC=
              mean(AUC)) 

FULL %>% 
  group_by(id,study) %>% 
  summarise(AUC=AUC(time,conc,method='trapezoid')) %>% 
  group_by(study)  %>% 
  summarise(mAUC=
              mean(AUC)) %>% 
  summarise(F=mAUC[study=="ORAL"]/mAUC[study=="IV"])

```

## Limitations of the linear trapezoid method 

```{r trapbest, echo=F, fig.cap="Participant 'A' IV concentration-time profile with curved line of best fit", fig.width=4.5,fig.height=2.5,fig.align = 'center'}
sIV<-subset(IV,id=="A"&time>4)
ggplot(data=sIV, aes(time,conc))+ 
    geom_ribbon(data=subset(sIV,  time >11 & time < 25), 
          aes(ymin=0,ymax=conc),
              fill='#82B446',
              alpha=0.2, outline.type = 'full',colour='gray44')+ 

  geom_point()+
  xlab("Time (hours)") + ylab("Drug concentration (mg/L)")+
  geom_line()+ 

  geom_line()  +
  geom_function(fun=function(x) (1000/27.75)*exp((-0.17)*x),linetype='twodash', col='red')+
  labs(caption = "Line of best fit shown in red")

```

It is worth noting the trapezoid method has limitations. Most notably that it will *overestimate* $AUC$ by virtue of taking a straight line between points, when the true concentration will fall in an logarithmic (curve) fashion (Figure \@ref(fig:trapbest)). There are methods to improve this accuracy, such as the logarithmic trapezoidal method. With this method, the 'trapezoid' is given a curved top (like the red line in Figure \@ref(fig:trapbest)). 


\newpage 

# Using <TT>R</TT> package <TT>PKNCA</TT> to undertake non-compartmental analysis  

If that all seems like too much hard work and <TT>R</TT> should just jolly well get on and do for us then you are right! There are a few non-compartmental analysis packages available, we will use <TT>PKNCA</TT>. Remember you will need to run <TT>install.packages('PKNCA')</TT> the first time you wish to use this package, once it has been installed you need to run <TT>library(PKNCA)</TT>. 


**How to:**

* Use <TT>PKNCA</TT> to undertake non-compartmental analysis on the oral teaching dataset

To run non-compartmental analysis using this package, we need to give <TT>PKNCA</TT> some dosing data. All the participants were given a dose of 1000 mg of the anti-epileptic at time $t=0$. To add a new column (variable) using <TT>\%>\%</TT> we use the <TT>mutate</TT> function.

```{r, results='hold'}

dORAL <- ORAL %>% # we will need to save this as a new object
  group_by(id) %>%
  slice(1) %>%
  mutate(dose = 1000) #this produces a new variable dose 1000

```

Now we can call and run <TT>PKNCA</TT>. First we create a sampling (or concentration) object and then a dosing object. In each function we instruct <TT>PKNCA</TT> of the names of our data object and the relevant variables in it. Then we combine the two and run our non-compartmental analysis.  

```{r, eval=FALSE, message=F}

conc<-PKNCAconc(ORAL, conc~time|id) #create sampling object
dose<-PKNCAdose(dORAL, dose~time|id) #create dosing object
full<-PKNCAdata(conc, dose)  #combine
results<-pk.nca(full)  # analyse
summary(results) #summarise the results
```
```{r, echo=FALSE}

conc<-PKNCAconc(ORAL, conc~time|id) #create sampling object
dose<-PKNCAdose(dORAL, dose~time|id) #create dosing object
full<-PKNCAdata(conc, dose)  #combine
results<-pk.nca(full)  # analyse
summary(results) #summarise the results

knitr::kable(summary(results), booktabs = TRUE, caption='PKNCA output') %>%
  kable_styling(bootstrap_options = "striped")
```

Look at the value for $AUC$ produced here and compare it to the one earlier. There is a discrepancy because by default, PKNCA uses a slightly different method to calculate $AUC$. 

# Clearance, half-life and volume of distribution

## Clearance ($Cl$)

Clearance describes the volume of plasma from which drug is completely removed per unit time. 

$$Cl = \frac{Dose}{AUC_{0-\infty}}$$

Note that this relationship holds exactly only for IV administration. For oral data, the calculated value reflects apparent clearance ($Cl/F$), since bioavailability ($F$) cannot be separated from clearance without an IV reference.

## Half-life ($t_{1/2}$)

Half-life is the time taken for drug concentration to fall by half. For a drug following one-compartment kinetics with first-order elimination, this is constant regardless of the starting concentration:

$$t_{1/2} = \frac{0.693 \times V_d}{Cl}$$
## Volume of distribution ($V_d$)

Volume of distribution describes how widely a drug distributes through the body. It is not a real anatomical volume, it is a proportionality constant relating the total amount of drug in the body to the measured plasma concentration.

For IV data, $V_d$ can be estimated by back-extrapolating the concentration-time curve to $t=0$ to obtain $C_0$ (concentration of drug when $t=0$):

$$V_d = \frac{Dose}{C_0}$$

For oral data this approach is not valid, as absorption is not instantaneous and $C_0$ cannot be meaningfully defined. Instead, only **apparent** volume of distribution ($V_d/F$) can be estimated from oral data, since bioavailability ($F$) cannot be separated from $V_d$ without an IV reference:

$$V_d/F = \frac{Cl/F \times t_{1/2}}{0.693}$$


\newpage

# Limitations of non-compartmental analysis

```{r traperror, echo=F, fig.cap="Concentration-time profile showing impact of reduced sampling on AUC calculation", fig.width=4.5,fig.height=2.5,fig.align = 'center'}

amox<- read.csv("https://raw.githubusercontent.com/dlonsdal/SGUL_PK_data/refs/heads/main/amoxicillinPK.csv")

Aamox<-amox%>%
        subset(ID=='A')

ggplot(data=Aamox, aes(time,conc))+ 
    geom_ribbon(data=subset(Aamox,time < 5&time!=2), 
          aes(ymin=0,ymax=conc),
              fill='#B4464B',
              alpha=0.1, outline.type = 'full',colour='gray44')+ 
    geom_ribbon(data=subset(Aamox,  time < 3), 
          aes(ymin=0,ymax=conc),
              fill='blue',
              alpha=0.1, outline.type = 'full',colour='gray44')+ 
    geom_ribbon(data=subset(Aamox,  time >1 & time < 5), 
          aes(ymin=0,ymax=conc),
              fill='#82B446',
              alpha=0.1, outline.type = 'full',colour='gray44')+
  geom_point()+
  theme_bw(base_size=9)+
  xlab("Time (hours)") + ylab("Concentration of amoxicillin (mg/L)")+
  theme_bw(base_size=9)+
  geom_line() +
  geom_function(fun=function(x) 1000/23.65*exp((-11.4/23.65)*x),col='red')+
  labs(caption = "Pink shaded area shows the increased error associated with reduced sampling \n (time = 2 sample removed)")

```

Non-compartmental analyses are useful for many reasons. For example, they use relatively straightforward algebra with few assumptions about the data to calculate summary pharmacokinetic information. Most regulators will require non-compartmental analysis of data as part of regulatory submissions. 

There are some disadvantages that are worth being aware of. These include:

* Need for rich sampling (multiple samples from each subject)
* Use of direct observation of $C_{max}$ and $T_{max}$ means *true* values are missed for any particular subject
* Increasing accuracy requires increased number of participants or samples, which can be technically challenging and expensive

Have a look at Figure \@ref(fig:traperror), where a situation is imagined where $time=2$ data point is not sampled. You can see the likely impact on the estimate of AUC from this. 