---
title: "Introduction to model diagnostics and covariate modelling"
output:
  bookdown::html_document2:
    fig_caption : TRUE
    number_sections: no
    toc: yes
    toc_float: yes
    
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{booktabs}
  - \usepackage{caption}
  - \captionsetup{labelfont=bf,margin=12pt}
urlcolor: blue
---

<style>
h1.title {
  font-size: 48px;

  text-align: center;
}
h1 {
  font-size: 24px;
}
h2 {
  font-size: 20px;
}
h3 {
  font-size: 16px;
}

h4.author { 
  text-align: center;
}
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
library(tidyverse)
library(kableExtra)
library(gridExtra)
library(PKNCA) 
library(RColorBrewer)
library(flextable)
library(nlmixr2)
library(gtsummary)
```

# Learning outcomes

* Explain what a covariate is 
* Give examples of patient characteristics that might influence pharmacokinetic parameters
* Interpret the results of a covariate analysis 
* Describe how a population pharmacokinetic model could be used to inform dosing decisions for individual patients or patient groups

# Tasks

* Fit a one-compartment IV bolus model for the combined dataset
* Test for covariate effect of weight on volume of distribution
* Assess whether including a covariate effect of weight on volume of distribution improves the model
* Test for covariate effect of sepsis on clearance
* Assess whether including a covariate effect of sepsis on clearance improves the model

## Extension tasks

* Test for and assess other covariate effects on clearance and volume of distribution. In the dataset, you have creatinine clearance, albumin, age...
* Assess whether changing to a different residual error model improves the fit 

# Packages you need for this session

```{r, eval=FALSE}
library(tidyverse)
library(nlmixr2)
```

# Dataset for tasks during this session

I have combined the two datasets you have been using throughout the day together. 

```{r datasets, eval=FALSE}

combined <- read_csv("https://raw.githubusercontent.com/c-swire/pkcourse/refs/heads/main/combined_ca.csv")

```

# Datasets used in workbook examples

In this Session the exemplars are:

* Sparse furosemide dataset used in Session 2
* Phenobarbital dataset that comes with the <TT>nlmixr2</TT> package

```{r dataset,warning=F,message=F}
sparse<- read_csv("https://raw.githubusercontent.com/dlonsdal/SGUL_PK_data/main/sparsefru.csv")

pheno <- pheno_sd
```

# Definition of a covariate

In statistical terms a variable is something that can be quantified in some way - e.g. measured or counted. 

A covariate is a variable that changes (varies) with (co) the outcome of interest. In modelling covariates are variables additional to the core predictor variable in the model (e.g. dose) that may be used to help model the outcome variable (e.g. concentration). 

An example might be that weight would be expected to affect volume of distribution for many drugs and so it might be reasonable to include a weight effect on the volume of distribution parameter in a pharmacokinetic model.  

Covariates can be continuous, categorical, discrete...

# Why bother to include covariates in a model?

We have discussed how non-linear mixed-effects modelling allows us to describe inter-individual differences in pharmacokinetic parameters - these are our $\eta$ terms. 

As an extension of this, it would be nice to try and understand and explain some of this variability so that we can:

* Explain it biologically, rather than just mathematically 
* Use this information to make decisions about our drugs
  + For example if we find that renal function is linked with clearance of a drug, we might recommend different doses for those with kidney disease

# Choosing covariates to include in a model

The choice of what covariates one should include in a model is complex. 

Options include:

* Test the effect of all variables in a data set on all parameters in the model. This can be a time-consuming (and processor hungry) process. 
* Knowledge about the drug's pharmacology (prior information) may guide the choice of which variables to test. E.g. if a drug is renally excreted, you might choose to test the effect of some measure of kidney function on drug clearance. 
* There are some tools that can help us. For example we can plot pharmacokinetic parameters (or $\eta$) against potential covariates to look for trends/relationships to test

# Introduction to model diagnostics 

There are a wealth of diagnostic plots that can be used to think about how well a model fits data from an experiment. We will look at a few key examples.

These approaches can be used to assess whether adding a covariate improves a model's fit.

You are likely to come across these plots and terms if you read any population pharmacokinetic modelling papers. We have explained them in a bit of detail so you should be able to interpret/understand these plots in future. 

## DV vs PRED and DV vs IPRED

The most straightforward diagnostic plot available is a plot of observations (dependent variable, DV) vs predicted concentrations from a model. There are two subsets of this type of plot. The first, DV vs PRED takes predictions based on population parameters (i.e. excluding $\eta$s) and the second is DV vs IPRED, which bases predictions on individual values for model parameters (including $\eta$s).

Figure \@ref(fig:frufit) shows these diagnostic plots for the furosemide data we used in Session 2.

```{r hide, eval=T, include=F, warning=F, message= F}

# First make the model
one.cmt <- function() {
  ini({ 
    tka <-log(1.3) # set the initial estimates for each parameter, note that ka will not be required for IV datasets
    tcl <- log(8) 
    tv <- log(20)
    eta.ka~0.1 # inter-individual variabilities, again note ka line this is not required for IV datasets
    eta.cl ~ 0.1
    eta.v ~ 0.1
    prop.sd <- 0.1 # residual variability
  })
  model({ # define the model
    ka <- exp(tka + eta.ka) # ka will not be required for IV datasets
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ prop(prop.sd) 
  })
}

# Fit the data using this model
fit<-nlmixr2(one.cmt,sparse,est="saem",
       list(print=0),
       table = tableControl(cwres = TRUE))
```

```{r frufit,echo=T,eval=T,fig.cap="Basic diagnostic plots for furosemide one-compartment, oral-absorption model fit from Chapter 4", fig.subcap=c('DV vs PRED','DV vs IPRED'),fig.ncol = 2, out.width = "50%", fig.align = 'center', warning=F,message=F}

modfit<-data.frame(fit) # make a dataframe out of the output of the model

ggplot(modfit,aes(x=DV,y=PRED))+
  geom_point()+
  theme_bw(base_size=18)+
  geom_abline(intercept = 0, slope=1)+
  geom_smooth(col='red',se=F)

ggplot(modfit,aes(x=DV,y=IPRED))+
  geom_point()+
  theme_bw(base_size=18)+
  geom_abline(intercept = 0, slope=1)+
  geom_smooth(col='red',se=F)

```

Take a look at Figure \@ref(fig:frufit)(a) first. A *perfect* model fit would have predictions that were identical to the observations from the experiment. All of the points in the scatterplot would fall in a straight line along the line of unity ($DV=PRED$). However, *real-world* data are never perfect, so what we are really looking for in this plot is even spread of points above and below the line of unity. Any groups of data points deviating away from this line of unity might make us suspicious that there is some underlying variability not being modelled. 

In our example in Figure \@ref(fig:frufit)(a) we can see that there is pretty reasonable fit, with even distribution of data points. A red 'line of best fit' tracks fairly evenly along the line of unity. This suggests our model fit is pretty good. 

Figure \@ref(fig:frufit)(b) shows DV vs IPRED. 
This is individualised predictions, taking into account random effect associated with an individual ($\eta$). Again, the perfect model fit has all lines along the line of unity. DV vs IPRED plots will generally always look a little better than DV vs PRED plots in this regard. However, it is unusual to get such a tight fit as we see in Figure \@ref(fig:frufit)(b) as usually there is additional unexplained variability in data. The example here looks relatively perfect because I simulated these data and I did not simulate complex variability!

## Conditionally-weighted residuals

When we fit a pharmacokinetic model to data, we can assess how well the model fits by examining the difference between each observed concentration and the concentration our model predicted. This difference is called a residual. 

However, raw residuals can be misleading because their size tends to scale with the magnitude of the observation: residuals around a high concentration early after dosing will naturally be larger than those around a low concentration at later time points, making it difficult to assess model fit consistently across the full concentration-time profile. 

Conditionally weighted residuals (CWRES) address this by standardising the residuals so that they are comparable across the full range of observations. 

If the model fits well, CWRES should be roughly normally distributed with a mean of zero and a standard deviation of one, meaning we would expect approximately 99.7% of values to fall between −3 and +3. 

When we plot CWRES against time, we are looking for an even, random scatter of points around zero. Any systematic trend, such as a U-shape or a consistent drift above or below zero, suggests that the model is not capturing something important in the data and may need to be refined.

<TT>nlmixr2</TT> will output <TT>CWRES</TT> through an additional <TT>tableControl()</TT> command.

```{r add cwres, eval=F}

# Fit the data using this model
fit<-nlmixr2(one.cmt,sparse,est="saem",
       list(print=0),
       table = tableControl(cwres = TRUE)) # you need to add this line when you are running the model to output a CWRES plot
```

```{r cwres,echo=T,eval=T,fig.cap="Conditionally weighted residuals plotted against time from dose - furosemide dataset",fig.width=6,fig.height=3,fig.align = 'center',warning=FALSE,message=FALSE}
ggplot(modfit,aes(x=tad,y=CWRES))+
  geom_point()+
  theme_bw(base_size=9)+
  xlab('Time after dose (h)')+
  geom_abline(intercept = 0, slope=0)+
  geom_smooth(col='red',se=F)+
  ylim(-3,3)
```

Figure \@ref(fig:cwres) is consistent with a well-fitted model. 

Figure \@ref(fig:2comp) is an example using a different dataset which is not well-fitted using a basic one-compartment model. 

```{r 2comp, echo=F, cache=F,warning=FALSE,message=FALSE, fig.cap="Conditionally weighted residuals plotted against time from dose - 2-compartment dataset",fig.width=6,fig.height=3,fig.align = 'center'}

one.cmt <- function() {
  ini({ 
    tcl <- log(0.01) 
    tv <- log(1)
        eta.cl ~ 0.1
    eta.v ~ 0.1
    prop.sd <- 0.1
  })
  model({
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ prop(prop.sd) 
  })
}

drugxsim<-read_csv("https://raw.githubusercontent.com/dlonsdal/SGUL_PK_data/main/drugxsim.csv")

drugxfit<-nlmixr2(one.cmt,drugxsim,est="saem",
       list(print=0),
       table = tableControl(cwres = TRUE))

modfit2<-data.frame(drugxfit)

ggplot(modfit2,aes(x=tad,y=CWRES))+
  geom_point()+
  theme_bw(base_size=9)+
  xlab('Time after dose (h)')+
  geom_abline(intercept = 0, slope=0)+
  geom_smooth(col='red',se=F)+
  ylim(-3,3)

```

## The Objective Function Value (OFV)
Objective function value is a measure of model fit. I do not propose that you learn in-depth how it is calculated or what its mathematics are. 

What is important for you to note is that the objective function can be used to compare two models. In general, a lower objective function value indicates a superior model fit.

## Visual Predictive Check (VPC)

The diagnostic plots we have used so far (DV vs PRED, DV vs IPRED, CWRES) assess model fit by examining residuals. A complementary and widely used approach is the **Visual Predictive Check (VPC)**.

The principle is straightforward: if our model is a good description of the data, then concentrations simulated from the model should look like the concentrations we actually observed. To produce a VPC, the fitted model is used to generate prediction intervals, which are then plotted alongside the observed data. 

```{r vpc, eval=TRUE, message=FALSE, warning=FALSE, fig.cap="Visual Predictive Check for one-compartment model fit of furosemide data", fig.width=6, fig.height=4, fig.align='center'}

vpcPlot(fit) +
  theme_bw(base_size = 11) +
  xlab("Time after dose (h)") +
  ylab("Furosemide concentration (mg/L)")
```

The VPC plot contains several elements. The **solid black line** represents the median of the observed data at each timepoint. The **dashed black lines** represent the 5th and 95th percentiles of the observed data, capturing the spread of concentrations across individuals. The **shaded blue areas** represent the model predictions — the darker central band shows the 95% confidence interval around the predicted median, and the lighter outer bands show the 95% confidence intervals around the predicted 5th and 95th percentiles. 

A well-fitting model will have the solid line running through the darker central band, and the dashed lines running through the lighter outer bands throughout the concentration-time profile. In this example, the observed lines track reasonably well through their corresponding prediction intervals, suggesting an adequate model fit.

# Modelling continuous covariates

Let's dive in to some practical work and look at the phenobarbital data that comes with the <TT>nlmixr2</TT> package. 

These data were published in 1985(!) and come from some clinical data from neonates given the drug. Neonates were given a loading dose IV injection followed by a maintenance 12 hourly IV dose. 

### How to:

* Explore the data with some summary tables and plots
* Fit a one-compartment model to these data and report the pharmacokinetic parameters
* Make some goodness-of-fit plots

```{r,echo = TRUE,eval=TRUE, message=FALSE}
pheno<-pheno_sd

head(pheno)
```

```{r,echo = TRUE,eval=TRUE, message=FALSE}
summary(pheno)
```


```{r pheno3,echo = F,eval=T, fig.cap="Phenobarbital concentration-time plot",fig.width=5,fig.height=2.5,fig.align='center',message=FALSE, fig.show='hold'}
pheno$ID<-as.factor(pheno$ID)
ggplot(subset(pheno,EVID==0), aes(x=TIME, y=DV))+
  geom_point(size=0.5)+
  theme_bw(base_size=9)
```

We've got an ID variable, <TT>WT</TT> which is hopefully easy to interpret as weight and <TT>APGR</TT> which is APGAR score (a measure of how unwell a neonate is when they are born). 

The other variables are <TT>nlmixr2</TT> default variables, as explained in Session 2's notes. 

```{r load,echo = TRUE,eval=TRUE, message=FALSE}

pheno %>% 
  group_by(ID) %>% 
  slice(1) %>% 
  ungroup() %>% 
  select("WT", "APGR") %>% 
  tbl_summary() %>% 
  as_kable_extra(booktabs=TRUE, caption="Summary of weight and APGR characteristics")%>%
  kable_styling(latex_options =c("striped"))
```

Table \@ref(tab:load) provides a summary of the two covariates. 

Next, we will run a baseline model with no covariates and do some model diagnostics plots. 

```{r, eval=T,cache=F, message=F,warning=FALSE}
one.cmt <- function() {
  ini({ 
    tcl <- log(0.01) 
    tv <- log(1)
    eta.cl ~ 0.1
    eta.v ~ 0.1
    prop.sd <- 0.1
  })
  model({
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    linCmt() ~ prop(prop.sd)
  })
}

run1<-nlmixr2(one.cmt,pheno,est="saem", 
              list(print=0),table = tableControl(cwres = TRUE, keep='WT'))
```


```{r nlmixfix,echo=F,warning=T,message=F}

kbl(run1$par.fixed ,booktabs = T, caption="Fixed effect parameter estimates from pheno analysis")%>%  kable_styling(latex_options =c("striped"))


```

```{r nlmixran,echo=F,warning=T,message=F}
kbl(signif(diag(run1$omegaR),2), 
    caption="Random effects parameter estimates",
    booktabs = T,
    col.names = c("Estimate")) %>%  
  kable_styling(latex_options = c("striped"), 
                full_width = FALSE)
```

You can see in Table \@ref(tab:nlmixran) that there is quite a bit of inter-individual variability. 

For completeness, we should plot our diagnostic plots.

```{r phenofit,echo=F,eval=T,fig.cap="Basic diagnostic plots for one-compartment model fit of pheno data", fig.subcap=c('DV vs PRED','DV vs IPRED', 'CWRES vs TAD'),fig.ncol = 3, out.width = "33%", fig.align = 'center', warning=F,message=F}

modfit<-data.frame(run1)

ggplot(modfit,aes(x=DV,y=PRED))+
  geom_point()+
  theme_bw(base_size=27)+
  geom_abline(intercept = 0, slope=1)+
  geom_smooth(col='red',se=F)

ggplot(modfit,aes(x=DV,y=IPRED))+
  geom_point()+
  theme_bw(base_size=27)+
  geom_abline(intercept = 0, slope=1)+
  geom_smooth(col='red',se=F)

ggplot(modfit,aes(x=tad,y=CWRES))+
  geom_point()+
  theme_bw(base_size=27)+
  xlab('Time after dose (h)')+
  geom_abline(intercept = 0, slope=0)+
  geom_smooth(col='red',se=F)+
  ylim(-3,3)
```

## Making a reference

When we include a continuous covariate such as weight in a population pharmacokinetic model, we need to think carefully about how we express it. We need to choose a reference value i.e. the value at which the covariate term has no effect on the parameter estimate. For example, with the continuous covariate weight a sensible choice is the median weight of the study population.

We do this by expressing weight as:

$$WT_{centred} = \log\left(\frac{WT}{WT_{median}}\right)$$

When an individual has exactly the median weight, $\log(WT/WT_{median}) = \log(1) = 0$, and the covariate term disappears entirely. This means our typical clearance estimate corresponds to a patient of median weight, which is clinically interpretable. For individuals above or below the median, the covariate term scales clearance up or down accordingly.

Without this step, the reference point would be a weight of zero, which is biologically meaningless.

### How to:

* Calculate median weight for phenobarbital dataset

```{r mw, message=F,warning=FALSE}

mw <- pheno %>% 
  group_by(ID) %>% 
  slice(1) %>% # take only one weight per patient
  ungroup() %>% 
  summarise(median(WT)) %>% 
  as.numeric

mw
```
## Running the model

```{r, eval=T,cache=F, message=F,warning=FALSE}

pheno$rWT<- log(pheno$WT/mw) # to create centred weight covariate (rWT)

run2.mod <- function() {
  ini({ 
    tcl <- log(0.007) 
    tv <- log(1.4)
    tw <- 1 # initial estimate of covariate effect of weight on clearance
    eta.cl ~ 0.1
    eta.v ~ 0.1
    prop.sd <- 0.1
  })
  model({
    cl <- exp(tcl + eta.cl +rWT*tw) 
    v <- exp(tv + eta.v)
    linCmt() ~ prop(prop.sd) 
  })
}

#nlmixr2(one.cmt)
run2<-nlmixr2(run2.mod,pheno,est="saem",
              list(print=0),table = tableControl(cwres = TRUE))
```

```{r nlmixfix2,echo=F,warning=T,message=F}

kbl(run2$par.fixed ,booktabs = T, caption="Fixed effect parameter estimates from pheno analysis run 2")%>%  kable_styling(latex_options =c("striped"))
```

```{r nlmixran2,echo=F,warning=T,message=F}
kbl(signif(diag(run2$omegaR),2), caption="Random effects parameter estimates run 2",booktabs = T,col.names = c(
                           "Estimate"))%>%  kable_styling(latex_options =c("striped"))
```

Compare the parameter estimates from Table \@ref(tab:nlmixfix) & \@ref(tab:nlmixran)  and Table \@ref(tab:nlmixfix2) & \@ref(tab:nlmixran2). 

### Goodness of fit plots for phenobarbital data after Run 2

```{r phenofit2,echo=F,eval=T,fig.cap="Basic diagnostic plots for one-compartment  model fit of pheno data with weight on clearance as a covariate", fig.subcap=c('DV vs PRED','DV vs IPRED', 'CWRES vs TAD'),fig.ncol = 3, out.width = "33%", fig.align = 'center', warning=F,message=F}


modfit2<-data.frame(run2)
ggplot(modfit2,aes(x=DV,y=PRED))+
  geom_point()+
  theme_bw(base_size=27)+
  geom_abline(intercept = 0, slope=1)+
  geom_smooth(col='red',se=F)
ggplot(modfit2,aes(x=DV,y=IPRED))+
  geom_point()+
  theme_bw(base_size=27)+
  geom_abline(intercept = 0, slope=1)+
  geom_smooth(col='red',se=F)

ggplot(run2,aes(x=tad,y=CWRES))+
  geom_point()+
  theme_bw(base_size=27)+
  xlab('Time after dose (h)')+
  geom_abline(intercept = 0, slope=0)+
  geom_smooth(col='red',se=F)
```

The PRED vs DV plot shows improved model fit with run2 (Figure \@ref(fig:phenofit2) (a)). Although there is still some room for improvement. 

## Using the Objective Function Value

If you add one parameter to your model and compare the objective function with and without that parameter, a drop in objective function value >3.84 suggests a statistically significant (p<0.05) improvement in model fit. 

Have a look at the objective function for the two model runs.

```{r,eval=F}
run1$OBJF
run2$OBJF
```
```{r,echo=F}
726.5
673.6
```

With this drop in OFV, we can be confident that adding weight as a covariate on clearance significantly improves model fit. 

# Modelling categorical covariates

The easiest example of modelling a categorical covariate is where you have two categories — a common example of this might be smoking/non-smoking. In a dataset, this should be coded with '0' (reference group, e.g. non-smoking) or '1'.

The full model for the covariate, including inter-individual variability, is (with smoking as an example):

$$CL_i = \theta_{CL} \times e^{\eta_{CL}} \times (1 + SM \times \theta_{smoke})$$

As $SM$ is smoker (1=yes, 0=no), you can see that for a non-smoker the covariate term drops out entirely (since $SM=0$, $SM \times \theta_{smoke} = 0$), and individual clearance reduces to:

$$CL_i = \theta_{CL} \times e^{\eta_{CL}}$$

You could this to your model block with this code:

```{r categorical, eval=F}

cl <- exp(tcl + eta.cl) * (1 + SM*tsm)

```

For a smoker ($SM=1$), clearance is scaled by $(1 + \theta_{smoke})$, where $\theta_{smoke}$ represents the proportional difference in clearance between smokers and non-smokers. A positive value of $\theta_{smoke}$ indicates higher clearance in smokers, and a negative value indicates lower clearance.

# Simulations

Building and refining a population pharmacokinetic model is not an end in itself. Once we have a well-fitted model that adequately describes drug exposure and accounts for key sources of variability we can use it for simulations and dose optimisation.

Having established a covariate relationship between sepsis and clearance, we can derive individual predicted clearance estimates for each patient and use these to simulate concentration-time profiles under alternative dosing regimens. In a septic patient with enhanced renal clearance, standard antibiotic dosing may fail to achieve pharmacokinetic/pharmacodynamic targets. This hypothesis can be explored using a simulation from a well-fitted population pharmacokinetic model.

In simulation, the parameter estimates and their associated variability are used to generate a large number of simulated individuals. The proportion of simulated patients achieving the pharmacokinetic/pharmacodynamic target under each dosing scenario can then be compared across regimens to identify the optimal dose and frequency of administration.

This model-informed precision dosing framework is increasingly used in clinical practice for drugs with a narrow therapeutic index, and underpins adaptive dosing strategies in therapeutic drug monitoring programmes. It is also central to drug development, allowing prospective evaluation of dose selection and sampling strategies before committing to large and expensive studies.